RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-11-27 manuscripts/manuscript.Rmd
Edits to and partial rewrite of the Introduction.
Overall, I tried to provide for better flow, less redudancy, and clearer
motivation preceding or at least quickly following all statements about
features and capabilities.

Though R provides the most extensive and widely used collection of
comparative phylogenetic methods tools, R users lack a convenient,
extensible, and semantically aware format to import, exchange and archive 
phylogenetic trees, traits, and metadata.
The rapid increase in phylogenetics data and research methods that make
use of such data has introduced many new challenges and opportunities 
for sharing, both between software environments and 
between research groups [@Drew_2013; @Stoltzfus_2012].
The ever expanding needs of the phylogenetics community are rapidly
out-pacing the capacity of current standards, which is resulting in a
radiation of different data representations and a loss of interoperability
at the very time when the growth of available data and methods has made
that interoperability most valuable.

For example, the currently popular NEXUS file format cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair).  This has led many developers to
modify NEXUS in different ways for different needs, with the unfortunate
result that NEXUS files generated by one program can be incompatible
with another [@Vos_2012]. Without a formal grammar, software based on 
NEXUS files may also make inconsistent assumptions about tokens, quoting,
or element lengths.  @Vos_2012 estimates as many as 15% of the NEXUS
files in the CIPRES portal contain unrecoverable but hard to diagnose errors.

NeXML was developed to addresses each of these problems and also open doors
to new opportunities in handling phylogenetic data. First, NeXML is defined
by a precise grammar that can be programmatically **validated**. This helps
ensure that any NeXML file can always be read without errors by any 
NeXML parsing software. Second, NeXML is **extensible**: a user can define
representations of new information (as we will illustrate) without breaking
this compatibility. Third and most importantly, NeXML is **semantically rich**.
The use of formal semantics in NeXML provides a representation of data that
is not only machine-readable, but can also permits machine reasoning about
the data provided (see @Parr2011 for an overview on semantic reasoning in 
evolutionary informatics). Instead of asking the computer to
return "the data on line #93", we can expect the computer to reason about
taxonomic concepts (as we illustrate in the supplement).

The R language [@R] has become an important environment for phylogenetics
research, with over 30 packages available including several with over 
1000 citations [@taskview]. `RNeXML` makes the large and rapidly growing
archive of richly annotated, NeXML-based phylogenetics data available to
the R user and this powerful ecosystem of R tools for phylogenetics.
Further, R packages and users frequently add their own additional 
information to the phylogenies they work with: annotations of species,
stochastic character maps, trait values, model estimates and parameter 
values. By leveraging the semantic and extensible nature of NeXML, `RNeXML` 
provides R users a platform independent data exchange format to capture,
further annotate, exchange, and preserve this information. With `RNeXML`,
R users and developers can both better contribute to and take advantage 
of the rapidly growing infrastructure of evolutionary informatics [@Parr2011].  
By capturing information already available in the R environment, together
with the appropriate choice of defaults, `RNeXML` can facilitate meeting
the rising bar for best practices in data sharing in phylogenetic 
research [@Cranston_2014] while at the same reducing the burden
on the user.

In this paper, we introduce the reader to the concepts and use of
RNeXML and related technologies. We begin by discussing methods to read
phylogenetic tree data and character trait data in the NeXML format into
R, and also how to write phylogenetic data and character trait data into
NeXML. The heart of this section focuses on tools for reading and writing
the kind of semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. Starting simple, we use examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end with an example that uses semantic metadata to
extend the NeXML standard to describe stochastic character mapping: a kind of
data annotation that is currently available in data structures
provided by existing R packages
but lacks a consistent, extensible export format. We illustrate how NeXML
files can be formally published to a data repository directly from R.
Finally, we provide a brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

This paper has been written as a dynamic document 
[@Xie_2013].  The reader may paste the code blocks
shown into their R terminal to follow along with the examples shown.

>>>>

The popular statistical and mathematical computing platform R [@R] arguably provides the most extensive and widely used collection of
comparative phylogenetic methods tools <!-- needs a citation to support it -->.
Yet, R users lack a convenient and extensible way to import, exchange and archive 
phylogenetic data, trait data, and their accompanying rich metadata in
a data exchange standard that comprehensively supports current data
exchange needs, is predictably machine-readable, and is forward compatible.
In response to the rapid increase in the variety of data
relevant to comparative analysis, and the rising data interoperability
needs of integrative comparative research approaches, such an exhcange standard,
called NeXML, has recently
been developed [@Vos_2012], and is beginning to be adopted. One of the
hurdles to broader adoption of any exchange standard is the
availability of easy-to-use programmtic interfaces (APIs) to the
format, in particular for developers of tools that consume or generate
relevant data, so that supporting the standard for reading input and
writing output data does not require additional effort, compared to,
say, using other but inferior exchange formats. Here we present such
an API in the form of an R package, called RNeXML. RNeXML aims to
provide easy programmatic access to reading and writing NeXML documents,
tailored for the kinds of use-cases that will be common for users and
developers of the wealth of evolutionary, and in particular
comparative analysis methods within the R ecosystem.

Increasingly many comparative research questions have come to exploit
the opportunities from, and thus rely on the ability to reuse and
integrate previously generated or published data and
metadata. However, the ever expanding needs of the evolutionary
research community are rapidly outpacing the capabilities of the
currently still most widely used exchange standards, which were all
developed a decade or more ago. This has resulted in a radiation of
different data representations and exchange standard "flavors" that
are no longer interoperable at the very time when the growth of
available data and methods has made that interoperability most
valuable. The problem is particularly acute in light of the challenges
for finding and sharing phylogenetic data
[@Drew_2013; @Stoltzfus_2012], in particular in a way that is not
subject to loss of most data and metadata semantics, or that is
amenable to predictable machine interpretation [@Cranston_2014]. For
example, the currently popular NEXUS file format [@Maddison_1997] cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair).  This has led to
modifications of NEXUS in different ways for different needs, with the unfortunate
result that NEXUS files generated by one program can be incompatible
with another [@Vos_2012]. Without a formal grammar, software based on 
NEXUS files may also make inconsistent assumptions about tokens, quoting,
or element lengths.  @Vos_2012 estimates that as many as 15% of the NEXUS
files in the CIPRES portal contain unrecoverable but hard to diagnose errors.

A detailed account of how the NeXML standard addresses these and other
relevant challenges can be found in @Vos_2012. In brief, NeXML was
designed with the following important properties. First, NeXML is defined
by a precise grammar that can be programmatically **validated**. This
ensures that any NeXML file can be expected to be parsed and read
without errors by any NeXML-compliant software, provided the file
passes validation. Second, NeXML is **extensible**: a user can define
representations of new, previously unanticipated information (as we
will illustrate) without violating its defining grammar. Third and
most importantly, NeXML is **semantically rich**.
The use of formal semantics in NeXML makes its data representation
not only machine-readable in a predictable manner, but also enables machine reasoning about
the data and metadata (see @Parr2011 for an overview on semantic reasoning in 
evolutionary informatics). Instead of asking the computer to
return "the data on line #93" <!-- this is a pretty obscure example, -->
<!-- and I'm not even clear what you mean here, so having a hard time -->
<!-- coming up with an improvement -->, we can expect the computer to reason about
taxonomic concepts (as we illustrate in the supplement).

The R language and platform [@R] has become a powerful ecosystem for
evolutionary analysis, with over 30 packages available alone for
comparative phylogenetics, several of which have accumulated over 
1000 citations [@taskview]. `RNeXML` makes the sizable amount of
richly annotated phylogenetic data already existing in NeXML format
available to R users and R's ecosystem, in a way, as we show here,
that is no more involved than reading or writing data from other
phylogenetic data formats. `RNeXML` also carries immediate benefits
for its users. For example, comparative analysis R packages and users
frequently add their own metadata annotations to the phylogenies they work with, including annotations of species,
stochastic character maps, trait values, model estimates and parameter 
values. `RNeXML` affords R the capability to harness machine-readable
semantics and an extensible metadata schema to capture, preserve, and
share these and other kinds of information, all through an API instead
of having to understand in detail the schema underlying the NeXML
standard. To assist users in meeting the rising bar for best practices
in data sharing in phylogenetic research [@Cranston_2014], `RNeXML`
captures metadata information from the R environment to the extent
possible, and applies reasonable defaults. Beyond these more immediate
benefits, `RNeXML` plugs the R ecosystem into the growing informatics
infrastructure of evolutionary informatics [@Parr2011], in which
interoperable data and machine-readable semantics play important
enabling roles.

The rest of this paper is organized as follows. First, we introduce
the basic use and concepts of RNeXML for reading and writing
phylogenetic trees and character trait data from and to NeXML. We then
focus on RNeXML's support for reading and writing the kind of
semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. We include examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end this section with a concrete example illustrating
how RNeXML facilitates harnessing the forward-compatibility property
of NeXML, by devising a serialization for stochastic character
mappings, a data annotation generated by existing R packages but
lacking a standard exchange format. Next we illustrate how RNeXML
promotes data archiving and sharing by providing commands for directly
publishing NeXML files to a data repository. We conclude the paper
with a brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

This paper has been written as a dynamic document in R Markddown
[@Xie_2013], which directly exercises all of the code blocks shown every time
the typeset version is regenerated, providing immediate feedback on
whether the code as shown is complete and continues to work. It also
means that the reader may paste the code blocks into their own R
terminal to follow along with the examples.
